# Association Networks

[Karpinets et al][assLink] recently published a paper on analyzing data with association networks. This looks rather interesting, but they made a claim in their work that they could generate better networks of associations by first calculating the association of an annotation with all other annotations (annotation profile), and then calculating the correlations between annotation profiles.

The claim is that they get information about indirect associations this way that they don't get by looking at direct co-occurrences. They claim that when they do a direct network, all they get is a dense hairball.

However, I was wondering what they would have gotten had they trimmed their direct associations based on the edge weights (degree of overlap), as proposed by [Merico et al][emLink] for their tool enrichment map.

```{r knitrsetup, echo=FALSE}
# sets output for inline results nicely
knit_hooks$set(inline = identity) 

# kills knitr if there is an error, therefore we don't waste time generating error messages
knit_hooks$set(error = function(x, options) stop(x)) 
```

```{r rsetup}
# these are extremely useful options for working in R
options(save.defaults=list(compress="xz"), stringsAsFactors=FALSE)
```

## Libraries required

I am going to use **RCytoscape** for the network visualization, and **graph** in **R** to work with the networks that we generate.

```{r libraries}
require(graph)
require(RCytoscape)
require(Biobase)
```


## Get data

The original data comes from the **GOLD** database, and is a series of attributes about bacterial species.

```{r getData}
gb <- read.table("anet-v1.0/GB.data", sep="\t", header=T, quote="", strip.white=T)
```

## Arrange data

For each group of annotations, get all of the unique attributes, and then determine which entries have which attributes, generating a list of associated entries for each annotation.

For each annotation column, split on comma, and then do a reverse

```{r arrangeAtts}
attCols <- names(gb[,seq(2, ncol(gb))])

attTypes <- vector("character", 0)
attValues <- vector("character", 0)
attEntries <- vector("list", 0)

annNames <- seq(1, nrow(gb))

for (iType in 1:length(attCols)){
	tmpType <- attCols[iType]
	tmpAnn <- gb[,tmpType]
	tmpAnn <- strsplit(tmpAnn, ",")
	names(tmpAnn) <- annNames
	tmpAnn <- reverseSplit(tmpAnn)
							 
	attTypes <- c(attTypes, rep(tmpType, length(tmpAnn)))
	attValues <- c(attValues, names(tmpAnn))
	attEntries <- c(attEntries, tmpAnn)
	
}
```

This has generated a list of the classes of annotations, the names of the annotations themselves, and a list with the entries that each annotation was an attribute of. Now we can calculate the similarity of each annotation with all of the other annotations based on which entries had a particular annotation as an attribute. 

We will store this data in an adjacency matrix, with the degree of overlap (calculated using a Jaccard similarity score) stored as the edge weight.


```{r adjMatrix}
nodeNames <- paste(attTypes, attValues, sep=":")
graphDat <- new("graphNEL", nodes=nodeNames, edgemode="directed") # this is pretty important!

graphDat <- initNodeAttribute(graphDat, attribute.name="attType",
															attribute.type="char", default.value="undefined")
nodeData(graphDat, nodeNames, "attType") <- attTypes

graphDat <- initNodeAttribute(graphDat, attribute.name="attName",
															attribute.type="char", default.value="undefined")
nodeData(graphDat, nodeNames, "attName") <- attValues

graphDat <- initNodeAttribute(graphDat, attribute.name="label",
															attribute.type="char", default.value="undefined")
nodeData(graphDat, nodeNames, "label") <- nodeNames

graphDat <- initEdgeAttribute(graphDat, attribute.name="weight",
															attribute.type="numeric", default.value=0)

nEnt <- length(attValues)

# set up the full set of comparisons that need to be done
allComp <- expand.grid(seq(1,nEnt),seq(1,nEnt))
allComp <- allComp[(allComp[,2] > allComp[,1]),] # keep only where second is greater than first
allComp <- as.matrix(allComp)

# now go through allComp, using them as indices into the matrix
jmAll <- sapply(seq(1,nrow(allComp)), function(x){
	doComp <- allComp[x,]
	a1 <- attEntries[[doComp[1]]]
	a2 <- attEntries[[doComp[2]]]
	
	jm <- (length(intersect(a1,a2))) / (length(union(a1,a2)))
	jm
})

notZero <- jmAll != 0

allComp <- allComp[notZero,]

jmAll <- jmAll[notZero]
fromEdge <- nodeNames[allComp[,1]]
toEdge <- nodeNames[allComp[,2]]

graphDat <- addEdge(fromEdge, toEdge, graphDat, jmAll)

```

Load up Cytoscape, and pass the graph over.

```{r passGraph}
jmCy <- CytoscapeWindow('jmGraph', graph=graphDat)
displayGraph(jmCy)
setLayoutProperties(jmCy, "force-directed", list(edge_attribute='weight'))
layoutNetwork(jmCy, "force-directed")
```

This does result in a massive hairball, so lets grab the edge weights, and delete some of the lower ones.

```{r hideEdges}
edgeDat <- getAllEdgeAttributes(jmCy)
minEdges <- (edgeDat$weight <= 0.05)
edgeNames <- paste(edgeDat$source[minEdges],' (unspecified) ',edgeDat$target[minEdges],sep='')
selectEdges(jmCy, edgeNames)
hideSelectedEdges(jmCy)
layoutNetwork(jmCy, "force-directed")
```